---
title: "STAY550-FINAL-PROJECT"
output: html_notebook
author: Shuyi Tan
editor_options: 
  chunk_output_type: inline
---

```{r Library,message = FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(nlme)
library(wgeesel)
library(gee)
library(mice)
library(lme4)
```

```{r Load Table}
data <- read.table("/Users/shuyitan/Desktop/Winter_Term_2/STAT550-FINAL/Data/data_dep.txt",sep = ' ')

# Select Variables
data <- data %>% dplyr::select(SN,NEW_GRP,month,DDS1,DDS4,GSI) %>%
          dplyr::rename(study_group = 'NEW_GRP',
                        gender = 'DDS1',
                        education = 'DDS4'
                        )

# Adjust variable type
data <- data %>% mutate(SN = as.factor(SN),
                study_group = as.factor(study_group),
                month = as.factor(month),
                gender = as.factor(gender))

str(data)
```


```{r Data Imputation,message=FALSE}
data_imp <- mice::mice(data,m=5,maxit=50,meth='pmm',seed=500)
data_imputed <- complete(data_imp,1)


data_treat_imp <- data_imputed %>% filter(study_group == '1' )
data_cont_imp <- data_imputed %>% filter(study_group == '2' )

data_treat_imp <-  data_treat_imp %>%
                      mutate(SN = as.character(SN),
                             month = as.numeric(as.character(month)))

data_cont_imp <-  data_cont_imp %>%
                      mutate(SN = as.character(SN),
                             month = as.numeric(as.character(month)))


data_imputed <-  data_imputed %>%
                      mutate(SN = as.character(SN),
                             month = as.numeric(as.character(month)))

```


# Exploratory Data Analysis

```{r}
data %>%
  ggplot(aes(x=GSI)) +
  geom_histogram()

data %>% arrange(GSI,desc = TRUE)
```


```{r Distribution of Categorical Variable}
# Gender
# 1  Male; 2  Female           
  
data_male <-  data %>% filter(gender == '1' )
length(unique(data_male$SN))

data_female <-  data %>% filter(gender == '2' )
length(unique(data_female$SN))

#  Study Group
data_treatment <- data %>% filter(study_group == '1' )
length(unique(data_treatment$SN))


data_control <- data %>% filter(study_group == '2' )
length(unique(data_control$SN))


# Education 
data %>% ggplot(aes(x=education)) + 
  geom_bar() + 
  scale_x_binned() + 
  labs(x = 'Education (in years)')
```

```{r missing rate at each time point}
data %>% group_by(month) %>%
  summarize(mis_count = sum(is.na(GSI))) %>% left_join(data %>% group_by(month) %>%
  tally(), by = 'month') %>%
  mutate(mis_prop = scales::percent(mis_count/n))

```

```{r}
# Remove record whose GSI is missing at all timep points or 4 time points
data_wide <- data %>% group_by(SN) %>%
  dplyr::select(SN,month,GSI) %>%
   pivot_wider(names_from = month, values_from = GSI) 

data_wide$NA_count <- rowSums(is.na(data_wide[,2:6]))


data <- data_wide %>% filter(NA_count < 4) %>% 
  dplyr::select(SN) %>% 
  inner_join(data, by = "SN")

```


```{r drop out time plotting}
data %>% group_by(month) %>% 
  ggplot(aes(x = GSI))  + 
  geom_histogram() +
  facet_wrap(~month)


data <- data %>% mutate(missing = ifelse(is.na(GSI), TRUE,FALSE))

chisq.test(data$missing,data$gender)
fisher.test(data$missing,data$gender)


data %>% group_by(missing) %>% 
  ggplot(aes(x = education))  + 
  geom_histogram() +
  facet_wrap(~missing)

chisq.test(data$missing,data$education)
fisher.test(data$missing,data$education,'simulate.p.value=TRUE')

```

```{r}
pd <- position_dodge(0.1) 


data_control %>% group_by(SN) %>%
  mutate(month = as.character(month),month = as.numeric(month)) %>%
  filter(is.na(GSI) == FALSE) %>%
  summarize(drop_out = max(month)) %>%
  right_join(data_control, by = 'SN') %>%
  mutate(month = as.factor(month),drop_out = as.factor(drop_out)) %>%
  group_by(drop_out,month) %>%
  filter(drop_out != '0') %>%
  summarize(mean_score = mean(GSI, na.rm = TRUE), sd_score = sd(GSI, na.rm = TRUE)) %>%
  filter(is.nan(mean_score) != TRUE) %>%
  ggplot(aes(x = month, y = mean_score, group=drop_out, color = drop_out)) +
  geom_point(position=pd) + 
  geom_line(position=pd) + 
  labs(y = 'Mean of GSI', x= 'Month') + 
  ggtitle('Control Group')
  #geom_errorbar(aes(ymin=mean_score-sd_score, ymax=mean_score+sd_score), width=.1,position=pd)


data_treatment %>% group_by(SN) %>%
  mutate(month = as.character(month),month = as.numeric(month)) %>%
  filter(is.na(GSI) == FALSE) %>%
  summarize(drop_out = max(month)) %>%
  right_join(data_treatment, by = 'SN') %>%
  mutate(month = as.factor(month),drop_out = as.factor(drop_out)) %>%
  group_by(drop_out,month) %>%
  filter(drop_out != '0') %>%
  summarize(mean_score = mean(GSI, na.rm = TRUE), sd_score = sd(GSI, na.rm = TRUE)) %>%
  filter(is.nan(mean_score) != TRUE) %>%
  ggplot(aes(x = month, y = mean_score, group=drop_out, color = drop_out)) +
  geom_point(position=pd) + 
  geom_line(position=pd) + 
  labs(y = 'Mean of GSI',x = 'Month') + 
  ggtitle('Intervention Group')



data %>% group_by(SN) %>%
  mutate(month = as.character(month),month = as.numeric(month)) %>%
  filter(is.na(GSI) == FALSE) %>%
  summarize(drop_out = max(month)) %>%
  right_join(data, by = 'SN') %>%
  mutate(month = as.factor(month),drop_out = as.factor(drop_out)) %>%
  group_by(drop_out,month) %>%
  filter(drop_out != '0') %>%
  summarize(mean_score = mean(GSI, na.rm = TRUE), sd_score = sd(GSI, na.rm = TRUE)) %>%
  filter(is.nan(mean_score) != TRUE) %>%
  ggplot(aes(x = month, y = mean_score, group=drop_out, color = drop_out)) +
  geom_point(position=pd) + 
  geom_line(position=pd) + 
  labs(y = 'Mean of GSI', x= 'Month') 
  
```

```{r all subjects box plot}
data %>% group_by(month) %>%
  ggplot(aes (x = month, y = GSI)) + 
  geom_boxplot() +
  labs(x = 'Month')

```




```{r randomly-selected 20 subjects}
data %>%
  filter(SN %in% sample(data$SN,20)) %>%
  ggplot(aes(x = month, y = GSI, group = SN)) +
  geom_point() +
  geom_line() + 
  labs(x = 'Month')
```


## Treatment group

```{r}
# Treatment group
p1 <- data_treatment %>% ggplot(aes(x = month, y = GSI, group = SN)) +
  geom_point() +
  geom_line()
```



```{r treatment box plot}
data_treatment %>% #group_by(month) %>%
  ggplot(aes(x = month, y= GSI)) +
  geom_boxplot()
```

```{r}
data_treatment %>% 
  group_by(month) %>%
  summarize(mean = mean(GSI, na.rm = TRUE),
            SD = sd(GSI, na.rm = TRUE))
```


```{r two-sample t-test in intervention group}
t.test(data_0_treat$GSI,data_3_treat$GSI)
t.test(data_3_treat$GSI,data_6_treat$GSI)
t.test(data_6_treat$GSI,data_18_treat$GSI)
t.test(data_18_treat$GSI,data_60_treat$GSI)



wilcox.test(data_0_treat$GSI,data_3_treat$GSI)
wilcox.test(data_3_treat$GSI,data_6_treat$GSI)
wilcox.test(data_6_treat$GSI,data_18_treat$GSI)
wilcox.test(data_18_treat$GSI,data_60_treat$GSI)

```

```{r}
t.test(data_0_cont$GSI,data_3_cont$GSI)
t.test(data_3_cont$GSI,data_6_cont$GSI)
t.test(data_6_cont$GSI,data_18_cont$GSI)
t.test(data_18_cont$GSI,data_60_cont$GSI)



wilcox.test(data_0_cont$GSI,data_3_cont$GSI)
wilcox.test(data_3_cont$GSI,data_6_cont$GSI)
wilcox.test(data_6_cont$GSI,data_18_cont$GSI)
wilcox.test(data_18_cont$GSI,data_60_cont$GSI)
```



## Control Group


```{r control box plot}
data_control %>% #group_by(month) %>%
  ggplot(aes(x = month, y= GSI)) +
  geom_boxplot()
```


## Treatment and Control group

```{r Group means over time}
# Group means over time
data %>% group_by(study_group, month) %>%
  summarize(mean_GSI = mean(GSI,na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = mean_GSI, group = study_group, color = study_group)) +
  geom_point() + 
  geom_line()

```

```{r cross sectional features at selected time, fig.width=4, fig.height=1.5}


data %>% group_by(month) %>%
  ggplot(aes(x=study_group, y=GSI, fill = study_group)) + 
  geom_boxplot() + 
  facet_wrap(~month, ncol = 5) + 
  labs(x = "Month") + 
  labs(fill = "Study Group") + 
  scale_fill_discrete(labels = c("1:Intervention", "2: Control")) + 
  ggtitle('Cross Sectional Effect of the Intervention')
```



```{r Deal with outliers, fig.width=4, fig.height=1.5}
Q <- quantile(data$GSI, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(data$GSI, na.rm = TRUE)
up <-  Q[2]+1.5*iqr # Upper Range  
low<- Q[1]-1.5*iqr # Lower Range

data_eliminated<- subset(data, data$GSI > (Q[1] - 1.5*iqr) & data$GSI < (Q[2]+1.5*iqr))


data_eliminated %>% group_by(month) %>%
  filter(month != '0') %>%
  ggplot(aes(x=study_group, y=GSI, fill = study_group)) + 
  geom_boxplot() + 
  facet_wrap(~month, ncol = 5) + 
  labs(x = "Month") + 
  labs(fill = "Study Group") + 
  scale_fill_discrete(labels = c("1:Intervention", "2: Control")) + 
  ggtitle('Cross Sectional Effect of the Intervention')

data <- data_eliminated
```

```{r descriptive table}
data_eliminated %>% 
  filter(month != '0') %>% 
  group_by(study_group,month) %>%
  summarize(mean = mean(GSI, na.rm = TRUE), sd = sd(GSI,na.rm = TRUE))
```


To confirm the effects of removal of device, two sample t-tests and Wilcoxon tests were conducted for each period. By checking the normality of the scores in each level, it appears that the data are likely to be normally distributed after removing extreme values/ outliers. Additionally, variance was checked to be fairly constant in each period.

```{r check normality}

data <- data_eliminated


data_0_treat <- data %>% filter(month == '0', study_group == '1')
data_0_cont <- data %>% filter(month == '0', study_group == '2')

data_3_treat <- data %>% filter(month == '3', study_group == '1')
data_3_cont <- data %>% filter(month == '3', study_group == '2')

data_6_treat <- data %>% filter(month == '6', study_group == '1')
data_6_cont <- data %>% filter(month == '6', study_group == '2')


data_18_treat <- data %>% filter(month == '18', study_group == '1')
data_18_cont <- data %>% filter(month == '18', study_group == '2')


data_60_treat <- data %>% filter(month == '60', study_group == '1')
data_60_cont <- data %>% filter(month == '60', study_group == '2')


# check normality 
shapiro.test(data_0_treat$GSI)
shapiro.test(data_0_cont$GSI)


shapiro.test(data_3_treat$GSI)
shapiro.test(data_3_cont$GSI)


shapiro.test(data_6_treat$GSI)
shapiro.test(data_6_cont$GSI)


shapiro.test(data_18_treat$GSI)
shapiro.test(data_18_cont$GSI)

shapiro.test(data_60_treat$GSI)
shapiro.test(data_60_cont$GSI)

# Normality all satisfied! 

```


```{r check variance}

data_3 <- data %>% filter(month == '3')
data_6 <- data %>% filter(month == '6')
data_18 <- data %>% filter(month == '18')
data_60 <- data %>% filter(month == '60')


var.test(GSI ~ study_group, data = data_3)

var.test(GSI ~ study_group, data = data_6)

var.test(GSI ~ study_group, data = data_18)

var.test(GSI ~ study_group, data = data_60)

```






```{r two sample t test}

t.test(data_3_treat$GSI,data_3_cont$GSI)

t.test(data_6_treat$GSI,data_6_cont$GSI)

t.test(data_18_treat$GSI,data_18_cont$GSI)

t.test(data_60_treat$GSI,data_60_cont$GSI)

```

```{r Wilcoxon test}
wilcox.test(data_3_treat$GSI,data_3_cont$GSI)
wilcox.test(data_6_treat$GSI,data_6_cont$GSI)
wilcox.test(data_18_treat$GSI,data_18_cont$GSI)
wilcox.test(data_60_treat$GSI,data_60_cont$GSI)
```

```{r Overall t test and Wilcoxon test}
t.test(data_treatment$GSI, data_control$GSI)
wilcox.test(data_treatment$GSI, data_control$GSI)
```



# Confirmative Analysis

## Treatment group
```{r determine random effect}
data_treatment <- data_treatment %>% mutate(SN = as.character(SN),
                          SN = as.numeric(SN),
                          month = as.character(month),
                          month = as.numeric(month)
                          )

data_treatment_lme <- groupedData(GSI ~ month|SN,data = data_treatment)
attach(data_treatment_lme)

plot(data_treatment_lme[1:100,])

fit_lm <- lmList(GSI~month|SN,data_treatment_lme,na.action=na.omit)
plot(intervals(fit_lm),axes=FALSE)
```

```{r compare random effect}
fit_ran_intercept <- lme(GSI~month,random = ~1|SN, data=data_treatment_lme)

fit_ran_both <- lme(GSI~month,random = ~month|SN, data=data_treatment_lme)

anova(fit_ran_intercept,fit_ran_both)
```

```{r check assumption of random-both-model }
# Q-Q plot for the random effects
qqnorm(fit_ran_both, ~ranef(.))

# QQ plot for residual
qqnorm(fit_ran_both, ~ resid(., type = "p"))


# Variance
plot(fit_ran_both)
#  It also shows relatively constant variance across the fitted range. The slight reduction in apparent variance on the right and left of the graph are likely a result of there being fewer observation in these predicted areas.
```

```{r Weighted GEE model, message=FALSE}
#wgee(model, data, id, family, corstr, scale = NULL, mismodel = NULL, maxit=200, tol=0.001)
fit_wgee_treat <- wgee(GSI~month, data=data_treatment, id=data_treatment$SN,family="gaussian",
            corstr="exchangeable",scale=NULL)
summary(fit_wgee_treat)
```

```{r imputed GEE model}

gee_treat <- gee(GSI~month,data = data_treat_imp,id = data_treat_imp$SN,corstr="exchangeable")

summary(gee_treat)  
```

## Control Group

```{r determine random effect}
data_control <- data_control %>% mutate(SN = as.character(SN),
                          SN = as.numeric(SN),
                          month = as.character(month),
                          month = as.numeric(month)
                          )

data_cont_lme <- groupedData(GSI ~ month|SN,data = data_control)
attach(data_cont_lme)


fit_lm_cont <- lmList(GSI~month|SN,data_cont_lme,na.action=na.omit)
plot(intervals(fit_lm_cont),axes=FALSE)

```

```{r compare random effect}
fit_ran_intercept_cont <- lme(GSI~month,random = ~1|SN, data=data_cont_lme)
fit_ran_both_cont <- lme(GSI~month,random = ~month|SN, data=data_cont_lme)
anova(fit_ran_intercept_cont,fit_ran_both_cont)
```

```{r check assumption of random-intercept-model }
# Q-Q plot for the random effects
qqnorm(fit_ran_intercept_cont, ~ranef(.))

# QQ plot for residual
qqnorm(fit_ran_intercept_cont, ~ resid(., type = "p"))

# Variance
plot(fit_ran_intercept_cont)
```

```{r summary of LEM (Control)}
summary(fit_ran_intercept_cont)
```



```{r Weighted GEE model, message=FALSE}
#wgee(model, data, id, family, corstr, scale = NULL, mismodel = NULL, maxit=200, tol=0.001)
fit_wgee_cont <- wgee(GSI~month, data=data_control, id=data_control$SN,family="gaussian",
            corstr="exchangeable",scale=NULL)
summary(fit_wgee_cont)
```

```{r imputed GEE model}
gee_cont <- gee(GSI~month,data = data_cont_imp,id = data_cont_imp$SN,corstr="exchangeable")

summary(gee_cont)  


pvalue_intecept  <- 2*min(pnorm(15.671587), 1-pnorm(15.671587))  
pvalue_month  <- 2*min(pnorm(-2.660117), 1-pnorm(-2.660117))  
pvalue_intecept
pvalue_month
```


# Overall, treatemnt acts as predictor

```{r Determine Random Effect}

data_all <- data %>%
  mutate(month = as.numeric(as.character(month)))#,
         #treatment = ifelse(study_group == '1', '1','0'),
        # treatment = as.factor(treatment))


lmm <- lmer(GSI ~ study_group + month + gender + education  + (1 | SN), data = data_all,
    REML = FALSE)


lmm_both <- lmer(GSI ~ study_group + month + gender + education  + (month | SN), data = data_all,
    REML = FALSE)

anova(lmm,lmm_both)

summary(lmm)
```

```{r}
lmm_test <- lme(GSI ~ study_group + month + gender + education,random=~1|SN,data=data_all)
summary(lmm_test)
```

```{r check assumption of random-intercept-model }
# Q-Q plot for the random effects
qqnorm(lmm_test, ~ranef(.))

# QQ plot for residual
qqnorm(lmm_test, ~ resid(., type = "p"))

# Variance
plot(lmm_test)
```



```{r Weighted Gee}
data_all$gender <- data_imputed$gender
data_all$education <- data_imputed$education
#data_all <- as.data.frame(data_all)

fit <- wgee(GSI ~ study_group+month+gender + education, data=data_all, id=data_all$SN, family="gaussian",
            corstr="exchangeable", scale=NULL)

summary(fit)
```

```{r Imputed GEE model}
data_imputed$study_group <- relevel(data_imputed$study_group,ref = '2')

gee_all <- gee(GSI~study_group + month + gender + education,data = data_imputed,id = data_imputed$SN,corstr="exchangeable")
summary(gee_all)
```


```{r Calculate p-value for GEE}
# Intercept
2*min(pnorm(7.2505246), 1-pnorm(7.2505246))  
# study_group1
2*min(pnorm(0.9361824), 1-pnorm(0.9361824))  
# month
2*min(pnorm(-6.3076754), 1-pnorm(-6.3076754))  

# gender2
2*min(pnorm(4.0806887), 1-pnorm(4.0806887))  

# education
2*min(pnorm(-3.9239191), 1-pnorm(-3.9239191))  

```












